\newpage
\linespread{1.5}

\begin{center}
	% 设置小二号字体和黑体
	{\heiti \zihao{-2} \textbf{中文题目}}
\end{center}

\begin{center}
	% 设置小四号宋体，文本加粗，且“摘要”两个字中间空两个空格
	 {\songti \zihao{4} \textbf{摘~  要}}
\end{center}

{
\zihao{-4}

随着科技的不断发展，人工智能在各个领域的应用越来越广泛。尤其是在医疗、金融和教育等行业，人工智能技术已经逐渐渗透并发挥着重要作用。许多企业和科研机构也开始加大对人工智能技术的投入，推动了这一领域的迅速发展。未来，人工智能有望带来更多创新，改变我们的工作和生活方式。

在人工智能的众多应用中，机器学习和深度学习无疑是最为重要的技术之一。通过大数据和算法的不断优化，机器学习能够使系统在没有明确编程的情况下，从数据中学习规律并做出预测。深度学习则进一步提升了模型的表现，尤其是在图像识别、语音处理和自然语言处理等方面取得了显著进展。

然而，随着人工智能技术的快速发展，也带来了许多挑战和问题。例如，如何避免算法的偏见，如何保障数据隐私和安全，如何平衡技术发展与伦理问题等，都是当前亟需解决的重要议题。未来，随着人工智能在各行各业的深入应用，我们需要更加关注其社会影响，确保技术为人类带来积极的改变。

{\songti \zihao{-4} \textbf{关键词：}}XXXXX；XXXXXX；XXXXXX
}

(\textcolor{red}{关键词3-5个, 用分号“；”隔开, 不能全部采用英文缩写})

\newpage

\begin{center}
    \zihao{-2}
    \textbf{English Title}
\end{center}

\begin{center}
    \zihao{4}
    \textbf{Abstract}
\end{center}

{
\zihao{-4}

In recent years, the field of artificial intelligence has made significant advancements, particularly in the areas of machine learning and deep learning. These technologies have revolutionized a wide range of industries, from healthcare to finance, enabling more efficient processes and better decision-making. As a result, there is a growing demand for skilled professionals who can harness the power of AI to solve real-world problems.

One of the most exciting developments in AI is the progress made in natural language processing (NLP). With the ability to understand, interpret, and generate human language, NLP has paved the way for chatbots, virtual assistants, and machine translation tools. These advancements have made it possible to interact with computers in more natural and intuitive ways, enhancing user experience and making technology more accessible to a wider audience.

Despite the tremendous potential of AI, there are still significant challenges that need to be addressed. Ethical considerations, such as bias in algorithms and the impact on employment, remain critical issues. Moreover, as AI systems become more complex, ensuring transparency and accountability in their decision-making processes is essential. Therefore, ongoing research and collaboration between experts from various fields are crucial to ensuring the responsible development and deployment of AI technologies.

\textbf{Keywords:} LaTeX; thesis; tutorial
}

(\textcolor{red}{英文标题和摘要由对应的中文标题和摘要翻译而来，关键词3-5个，用分号“；”分隔})

